Exact expressions for double descent and implicit regularization \\ via surrogate random design

### 2022-08-23 design/continuous_nips/continuous.tex
Commit original repos

\newif\ifisarxiv
\isarxivtrue

@@

\title{Exact expressions for double descent
and implicit regularization
\\
via surrogate random design}
  \author{%
          Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \and
  Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \and
   Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
}

@@

\else
\documentclass[anon]{colt2020}

\title[Exact expressions for double descent and implicit
regularization]{Exact expressions for double descent and implicit
  regularization\\
  via surrogate random design}
\usepackage{times}

\coltauthor{%
  \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
  \addr Department of Statistics\\
  University of California, Berkeley\\
  \AND
  \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
  \addr Department of Statistics\\
  University of California, Berkeley\\
  \AND
  \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
  \addr ICSI and Department of Statistics\\
  University of California, Berkeley
}

\usepackage{float}
\usepackage{wrapfig}

@@

@@

\begin{abstract}
  Double descent refers to the phase transition that is exhibited by
  the generalization error of unregularized learning models when varying the ratio
  between the number of parameters and the number of training
  samples. The recent success of highly over-parameterized machine learning
  models such as deep neural networks has motivated a theoretical analysis of
  the double descent phenomenon in classical models such as linear
  regression which can also generalize well in the over-parameterized
  regime. We build on recent advances in Randomized Numerical Linear
  Algebra (RandNLA) to provide the first exact non-asymptotic
  expressions for double descent of the minimum norm linear
  estimator. Our approach involves constructing
  what we call a surrogate random design to replace the standard
  i.i.d.~design of the training sample. This surrogate design admits
  exact expressions for the mean squared error of the estimator while
  preserving the key properties of the standard design.
  We also establish an exact implicit regularization result for
  over-parameterized training samples. In particular, we show that, for
  the surrogate design, the implicit bias of the unregularized minimum
  norm estimator precisely corresponds to solving a ridge-regularized
  least squares problem on the population distribution.
\end{abstract}

\ifisarxiv\else
\begin{keywords}
  Least squares, determinantal point process,
  double descent, implicit regularization
\end{keywords}
\fi

@@

We build on methods from Randomized Numerical Linear Algebra (RandNLA) in order
to obtain \emph{exact non-asymptotic expressions} for the mean squared error
(MSE) of the Moore-Penrose estimator (see Theorem~\ref{t:mse}).  This provides
a precise characterization of the double descent phenomenon for perhaps the
simplest and most ubiquitous regression problem.  In obtaining these results,
we are able to provide precise formulas for the \emph{implicit regularization}
induced by minimum norm solutions of under-determined training samples,
relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}).
This result has been observed empirically for RandNLA
methods~\citep{Mah-mat-rev_JRNL}, but it has also been shown in deep
learning~\citep{Ney17_TR} and machine learning~\citep{Mah12} more generally.  To
obtain our precise results, we use a somewhat non-standard random design, which
we term surrogate random design (see Section~\ref{s:determinantal} for a
detailed discussion), and which we expect will be of more general interest.
Informally, the goal of a surrogate random design is to modify an original
design to capture its main properties while being ``nicer'' in some useful way.
In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
theoretically and empirically, that our surrogate design accurately preserves
the key properties of the original design when the data distribution is a
multivariate Gaussian.

@@

As the performance metric in our analysis, we use the \emph{mean
  squared error} (MSE), defined as
$\mathrm{MSE}[\wbh]=\E\big[\|\wbh-\w^*\|^2\big]$, where $\w^*$ is a fixed
underlying linear model of the responses.
In analyzing the MSE, we make the following standard assumption that
the response noise is homoscedastic with variance $\sigma^2$.
\begin{assumption}[Homoscedastic noise]\label{a:linear}
  Responses are $y(\x) = \x^\top\w^*+\xi$ where
$\xi\sim\Nc(0,\sigma^2)$.
\end{assumption}

\noindent
Our main result provides an exact expression for the MSE of the
Moore-Penrose estimator under our surrogate design denoted $\Xb\sim S_\mu^n$, where
$\mu$ is the $d$-variate distribution of the row vector $\x^\top$ and $n$ is the sample
size (details in Section~\ref{s:determinantal}). This surrogate is
used in place of the standard $n\times d$ random design $\X\sim\mu^n$, where $n$
data points (the rows of $\X$) are sampled independently from
$\mu$. Unlike for the standard design, our MSE formula is
fully expressible as a function of the covariance
matrix $\Sigmab_\mu=\E_\mu[\x\x^\top]$. To state our main result, we
need an additional minor assumption on $\mu$ which is satisfied by
most standard continuous distributions such as any multivariate
Gaussian with positive definite covariance matrix.
\begin{assumption}[General position]\label{a:general-position}
For $1\leq n \leq d$, if $\X\sim\mu^n$, then $\rank(\X)=n$ almost surely.
\end{assumption}

@@

\begin{theorem}[Exact non-asymptotic MSE]
\label{t:mse}
  If the response noise is homoscedastic % with variance~$\sigma^2$
  (Assumption~\ref{a:linear}) and $\mu$ is in
  general position (Assumption~\ref{a:general-position}), then for
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and
  $\yb_i=y(\xbb_i)$, %we have
  \begin{align*}
    \MSE{\Xb^\dagger\ybb} =
    \begin{cases}
    \sigma^2\,\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)\cdot
    \frac{1-\alpha_n}{d-n}\ +\
\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
{\tr((\Sigmab_\mu+\lambda_n\I)^{-1})}\cdot (d-n),
&\text{for }n<d,\\
\sigma^2\, \tr(\Sigmab_\mu^{-1}),& \text{for }n=d,\\
\sigma^2\,\tr(\Sigmab_\mu^{-1})\cdot\frac{1-\beta_n}{n-d},&\text{for
}n>d,
\end{cases}
  \end{align*}
with $\lambda_n\geq 0$ defined by
  $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$, \
  $\alpha_n=\det(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$
  and $\beta_n=\ee^{d-n}$.
\end{theorem}
\begin{definition}
  We will use $\Mc(\Sigmab_\mu, \w^*,\sigma^2,n)$ to denote the above expressions
  for $\MSE{\Xb^\dagger\ybb}$.
\end{definition}

\begin{figure}[t]
\centering
\subfigure[%
  Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
  numerical estimates even for non-isotropic
  features. Eigenvalue decay leads to a steeper
  descent curve in the under-determined regime ($n<d$).]{%
    \includegraphics[width=0.48\textwidth]{figs/descent-intro}
  }
\hfill
\subfigure[%
    The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.]{%
      \includegraphics[width=0.48\textwidth]{figs/descent-shrinkage}
    }
\caption{Illustration of the main results for $d=100$ and
$\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
eigenvalues decaying exponentially and scaled so that
$\tr(\Sigmab^{-1})=d$. We use our surrogate
formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
\ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
lines), accompanied by the empirical estimates based on the standard
i.i.d.~design (error bars are three times the standard error of the
mean). We consider three different condition numbers $\kappa$ of
$\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
$\w^*=\frac1{\sqrt{d}}\one$.}
\vspace{-5mm}
\label{f:intro}
\end{figure}

\noindent
Proof of Theorem \ref{t:mse} is given in Appendix \ref{a:mse-proof}.
For illustration, we plot the MSE expressions in Figure~\ref{f:intro}a,
comparing them with empirical estimates of the true MSE under the
i.i.d.~design for a multivariate Gaussian distribution
$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
$100$ and vary the number of samples $n$, observing a double descent
peak at $n=d$. We observe that our theory aligns well with
the empirical estimates, whereas
previously, no such theory was available except for special
cases such as $\Sigmab=\I$ (more details in Theorem \ref{t:asymptotic}
and Appendices \ref{sec:proof-of-t-asymptotic} and \ref{sec:asymp-conj-details}). The plots
show that varying the spectral decay of $\Sigmab$ has a significant effect on the
shape of the curve in the under-determined regime. We use the
horizontal line to denote the MSE of the null estimator
$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
decay rapidly, then the Moore-Penrose estimator suffers less error
than the null estimator for some values of $n<d$, and the curve
exhibits a local optimum in this regime.

One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$.
This expression is precisely the classical notion of \emph{effective
  dimension} for ridge regression regularized with
$\lambda_n$~\citep{ridge-leverage-scores}, and it arises here even though there is
no explicit ridge regularization in the problem being considered in
Theorem~\ref{t:mse}. 
The global solution to the ridge regression task (i.e., $\ell_2$-regularized
least squares) with parameter $\lambda$ is defined as:
\begin{align*}
\argmin_\w \Big\{\E_{\mu,y}\big[\big(\x^\top\w-y(\x)\big)^2\big]
    + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
  \lambda\I)^{-1}\v_{\mu,y},\quad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
\end{align*}
When Assumption \ref{a:linear} holds, then
$\v_{\mu,y}=\Sigmab_\mu\w^*$, however ridge-regularized least squares
is well-defined for much more general response models.
Our second result makes a direct connection between the (expectation
of the) unregularized minimum norm solution on the sample
and the global ridge-regularized solution.
While the under-determined regime (i.e., $n<d$) is of primary interest to us,
for completeness we state this result for arbitrary values of $n$ and $d$.
Note that, just like the definition of regularized least squares, this
theorem applies more generally than Theorem~\ref{t:mse}, in that it
does \emph{not} require the responses to follow any linear model as in
Assumption~\ref{a:linear} (proof in Appendix~\ref{s:unbiased-proof}). 
\begin{theorem}[Implicit regularization of Moore-Penrose estimator]
\label{t:unbiased}
For $\mu$ satisfying\footnote{The proof of Theorem \ref{t:unbiased}
  can be easily extended to probability measures $\mu$ that do not
  satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
  assumption here to simplify the presentation.} Assumption~\ref{a:general-position} and
  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined, if
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$, then
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big] =
    \begin{cases}
       (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} &\text{for }n<d,\\
        \Sigmab_\mu^{-1}\v_{\mu,y} &\text{for }n \ge d,
    \end{cases}
  \end{align*}
  where, as in Theorem \ref{t:mse}, $\lambda_n$ is such that the effective dimension
  $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
\end{theorem}

@@

Our exact non-asymptotic expressions in Theorem~\ref{t:mse} and
our exact implicit regularization results in Theorem~\ref{t:unbiased}
are derived for the surrogate design, but Figure \ref{f:intro}
suggests that they accurately describe the MSE (up to lower order
terms) also under the standard i.i.d.~design $\X\sim\mu^n$,
particularly when $\mu$ is a multivariate Gaussian.
As a third result, we can verify this in the cases where there exist
known expressions for the MSE under the i.i.d.~design (standard
Gaussian for the under-determined setting, and
arbitrary Gaussian for the over-determined one).
We give the proof in Appendix~\ref{sec:proof-of-t-asymptotic}.
\begin{theorem}[Asymptotic consistency of surrogate design]
\label{t:asymptotic}
Let $\rho=n/d\neq 1$, $\X\sim \mu^n$
and $y_i=y(\x_i)$ satisfy Assumption~\ref{a:linear}. If $d\geq
c_\rho=\frac3{|1-\rho|}$ and
\begin{align*}
  \mu &= \begin{cases}
    \Nc(\zero,\I)&\text{when }n<d-1,\\
    \Nc(\zero,\Sigmab),\ \Sigmab\succ\zero &\text{when }n>d+1,
  \end{cases}
\end{align*}
then the absolute difference between surrogate
expressions and the true MSE is bounded as follows:
% &\text{if}\qquad  (1)\quad \rho<1\ \text{ and }\ \mu=\Nc(\zero,\I)\qquad\text{or}\qquad
%     (2)\quad \rho > 1\ \text{ and }\ \mu=\Nc(\zero,\Sigmab),\
%     \Sigmab\succ\zero,\\
\begin{align*}
\Big|\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)\Big|\ \leq\
\frac{c_\rho}{d} \cdot \Mc(\Sigmab,\w^*,\sigma^2,n).
\end{align*}
\end{theorem}
\begin{remark}
For $n$ equal to $d-1$, $d$ or $d+1$, the true MSE under Gaussian
random design can be infinite, whereas the
surrogate MSE is finite and has a closed form expression.
\end{remark}

\noindent
Empirical estimates given in Figure
\ref{f:intro} suggest that the consistency of surrogate expressions
holds much more generally than it is stated above. Based on a detailed
empirical analysis described in Appendix \ref{sec:asymp-conj-details},
we conjecture that an asymptotic consistency 
result of the form similar to the statement of Theorem~\ref{t:asymptotic}
holds true in the under-determined regime without the assumption that
$\Sigmab=\I$ (see Conjectures \ref{c:wishart} and
\ref{c:projection}). In this case, no formula is known for 
$\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
Gaussian design naturally extend.

\subsection{Key techniques: surrogate designs and
  determinant preserving matrices}

>>>>

 \newif\ifisarxiv
 \isarxivfalse

@@

 \title{Exact expressions for double descent
 and implicit regularization
 \\
 via surrogate random design}
   \author{%
           \textbf{Micha{\l } Derezi\'{n}ski} \\
   Department of Statistics\\
   University of California, Berkeley\\
   \texttt{mderezin@berkeley.edu}\\
   \and
   \textbf{Feynman Liang} \\
   Department of Statistics\\
   University of California, Berkeley\\
   \texttt{feynman@berkeley.edu}
   \and
    \textbf{Michael W. Mahoney}\\
   ICSI and Department of Statistics\\
   University of California, Berkeley\\
   \texttt{mmahoney@stat.berkeley.edu}
 }

@@

\else

\documentclass{article}
\usepackage[final]{neurips_2020}

\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{amsmath}
\usepackage{cancel}
\usepackage{cleveref}
\usepackage{subfigure}
\usepackage{todonotes}
 \usepackage{wrapfig}
\usepackage{xcolor}
\usepackage{xfrac}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}

\title{Exact expressions for double descent and implicit regularization via
surrogate random design}
  \author{%
Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \And
Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
}
% \usepackage{times}

% \coltauthor{%
%   \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
%   \addr Department of Statistics\\
%   University of California, Berkeley\\
%   \AND
%   \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
%   \addr Department of Statistics\\
%   University of California, Berkeley\\
%   \AND
%   \Name{Zhenyu Liao} \Email{TODO}\\
%   \addr Department of Statistics\\
%   University of California, Berkeley\\
%   \AND
%   \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
%   \addr ICSI and Department of Statistics\\
%   University of California, Berkeley
% }

% \usepackage{float}
% \usepackage{wrapfig}

@@

\newcommand{\RED}{\color[rgb]{0.70,0,0}}

@@

\begin{abstract}
  Double descent refers to the phase transition that is exhibited by
  the generalization error of unregularized learning models when varying the ratio
  between the number of parameters and the number of training
  samples. The recent success of highly over-parameterized machine learning
  models such as deep neural networks has motivated a theoretical analysis of
  the double descent phenomenon in classical models such as linear
  regression which can also generalize well in the over-parameterized
  regime. We provide the first exact non-asymptotic
  expressions for double descent of the minimum norm linear
  estimator. Our approach involves constructing a special
  determinantal point process  which we call surrogate random
  design, to replace the standard i.i.d.~design of the training
  sample. This surrogate design admits exact expressions for the mean
  squared error of the estimator while preserving the key properties
  of the standard design. We also establish an exact implicit
  regularization result for over-parameterized training samples. In
  particular, we show that, for the surrogate design, the implicit bias
  of the unregularized minimum norm estimator precisely corresponds to
  solving a ridge-regularized least squares problem on the population
  distribution. In our analysis we introduce a new mathematical tool of
  independent interest: the class of random matrices for which
  determinant commutes with expectation.
\end{abstract}

@@

We build on methods from Randomized Numerical Linear Algebra (RandNLA) in order
to obtain \emph{exact non-asymptotic expressions} for the mean squared error
(MSE) of the Moore-Penrose estimator (see Theorem~\ref{t:mse}).  This provides
a precise characterization of the double descent phenomenon for the
linear regression problem. In obtaining these results,
we are able to provide precise formulas for the \emph{implicit regularization}
induced by minimum norm solutions of under-determined training samples,
relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}).
% This result has been observed empirically for RandNLA
% methods~\citep{Mah-mat-rev_JRNL}, but it has also been shown in deep
% learning~\citep{Ney17_TR} and machine learning~\citep{Mah12} more generally.
To obtain our precise results, we use a somewhat non-standard random
design, based on a specially chosen determinantal point process (DPP), which we
term surrogate random design. DPPs are a family of non-i.i.d.~sampling
distributions which are typically used to induce diversity in the
produced samples \citep{dpp-ml}. Our aim in using a DPP as a surrogate
design is very different: namely, to make certain quantities (such as the MSE)
analytically tractable, while accurately \emph{preserving} the underlying
properties of the original data distribution. This strategy might seem
counter-intuitive since DPPs are typically found most useful when they
\emph{differ} from the data distribution. However, we show both theoretically
(Theorem~\ref{t:asymptotic}) and empirically
(Section~\ref{sec:asymp-conj-details}), that for many commonly studied data
distributions, such as multivariate Gaussians, our DPP-based surrogate
design accurately preserves the key properties of the standard i.i.d.~design
(such as the MSE), and even matches it exactly in the high-dimensional
asymptotic limit. In our analysis of the surrogate design, we
introduce the concept of \emph{determinant preserving
  random matrices} (Section \ref{s:dp}), a class of random matrices for which determinant
commutes with expectation, which should be of independent interest.

%  To
% obtain our precise results, we use a somewhat non-standard random design, which
% we term surrogate random design (see Section~\ref{s:determinantal} for a
% detailed discussion), and which we expect will be of more general interest.
% Informally, the goal of a surrogate random design is to modify an original
% design to capture its main properties while being ``nicer'' in some useful way.
% In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
% theoretically and empirically, that our surrogate design accurately preserves
% the key properties of the original design when the data distribution is a
% multivariate Gaussian.

@@

As the performance metric in our analysis, we use the \emph{mean
  squared error} (MSE), defined as
$\mathrm{MSE}[\wbh]=\E\big[\|\wbh-\w^*\|^2\big]$, where $\w^*$ is a fixed
underlying linear model of the responses.
In analyzing the MSE, we make the following standard assumption that
the response noise is homoscedastic.
\begin{assumption}[Homoscedastic noise]\label{a:linear}
  The noise $\xi =y(\x)- \x^\top\w^*$ has mean $0$ and variance $\sigma^2$.
\end{assumption}

\noindent
Our main result provides an exact expression for the MSE of the
Moore-Penrose estimator under our surrogate design denoted $\Xb\sim S_\mu^n$, where
$\mu$ is the $d$-variate distribution of the row vector $\x^\top$ and $n$ is the sample
size. This surrogate is
used in place of the standard $n\times d$ random design $\X\sim\mu^n$, where $n$
data points (the rows of $\X$) are sampled independently from
$\mu$. We form the surrogate by constructing a determinantal point
process with $\mu$ as the background measure, so that $S_\mu^n(\X)\propto
  \pdet(\X\X^\top)\mu(\X)$, where $\pdet(\cdot)$ denotes the pseudo-determinant
(details in Section~\ref{s:determinantal}).  Unlike for the standard
design, our MSE formula is fully expressible as a function of the covariance
matrix $\Sigmab_\mu=\E_\mu[\x\x^\top]$. To state our main result, we
need an additional minor assumption on $\mu$ which is satisfied by
most standard continuous distributions (e.g., multivariate Gaussians).
\begin{assumption}[General position]\label{a:general-position}
  For $1\leq n \leq d$, if $\X\sim\mu^n$, then $\rank(\X)=n$ almost surely.
\end{assumption}

@@

\begin{theorem}[Exact non-asymptotic MSE]
  \label{t:mse}
  If the response noise is homoscedastic % with variance~$\sigma^2$
  (Assumption~\ref{a:linear}) and $\mu$ is in
  general position (Assumption~\ref{a:general-position}), then for
  any $\w \in \mathbb{R}^d$,
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}), and
  $\yb_i=y(\xbb_i)$, %we have
  \vspace{-2mm}
  \begin{align*}
    \MSE{\Xb^\dagger\ybb} =
    \begin{cases}
      \sigma^2\,\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)\cdot
      \frac{1-\alpha_n}{d-n}\ +\
      \frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
      {\tr((\Sigmab_\mu+\lambda_n\I)^{-1})}\cdot (d-n),
                                                                 & \text{for }n<d, \\
      \sigma^2\, \tr(\Sigmab_\mu^{-1}),                          & \text{for }n=d, \\
      \sigma^2\,\tr(\Sigmab_\mu^{-1})\cdot\frac{1-\beta_n}{n-d}, & \text{for
      }n>d,
    \end{cases}
  \end{align*}
  with $\lambda_n\geq 0$ defined by
  $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$, \
  $\alpha_n=\det(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$
  and $\beta_n=\ee^{d-n}$.
\end{theorem}
\begin{definition}\label{d:expression}
  We will use $\Mc = \Mc(\Sigmab_\mu, \w^*,\sigma^2,n)$ to denote the above expressions
  for $\MSE{\Xb^\dagger\ybb}$.
\end{definition}

\begin{figure}[t]
  \centering
  \subfigure[%
    Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
    numerical estimates even for non-isotropic
    features. Eigenvalue decay leads to a steeper
    descent curve in the under-determined regime ($n<d$).]{%
    \includegraphics[width=0.48\textwidth]{figs/descent-intro}
  }
  \hfill
  \subfigure[%
    The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.]{%
    \includegraphics[width=0.48\textwidth]{figs/descent-shrinkage}
  }
  \caption{Illustration of the main results for $d=100$ and
    $\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
    eigenvalues decaying exponentially and scaled so that
    $\tr(\Sigmab^{-1})=d$. We use our surrogate
    formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
    \ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
    lines), accompanied by the empirical estimates based on the standard
    i.i.d.~design (error bars are three times the standard error of the
    mean). We consider three different condition numbers $\kappa$ of
    $\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
    i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
    $\w^*=\frac1{\sqrt{d}}\one$.}
  %\vspace{-5mm}
  \label{f:intro}
\end{figure}

\noindent
Proof of Theorem \ref{t:mse} is given in Appendix \ref{a:mse-proof}.
For illustration, we plot the MSE expressions in Figure~\ref{f:intro}a,
comparing them with empirical estimates of the true MSE under the
i.i.d.~design for a multivariate Gaussian distribution
$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
$100$ and vary the number of samples $n$, observing a double descent
peak at $n=d$. We observe that our theory aligns well with
the empirical estimates, whereas
previously, no such theory was available except for special
cases such as $\Sigmab=\I$ (more details in Theorem~\ref{t:asymptotic}
and Section~\ref{sec:asymp-conj-details}). The plots
show that varying the spectral decay of $\Sigmab$ has a significant effect on the
shape of the curve in the under-determined regime. We use the
horizontal line to denote the MSE of the null estimator
$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
decay rapidly, then the Moore-Penrose estimator suffers less error
than the null estimator for some values of $n<d$, and the curve
exhibits a local optimum in this regime.

One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$.
This expression is precisely the classical notion of \emph{effective
  dimension} for ridge regression regularized with
$\lambda_n$~\citep{ridge-leverage-scores}, and it arises here even though there is
no explicit ridge regularization in the problem being considered in
Theorem~\ref{t:mse}.
The global solution to the ridge regression task (i.e., $\ell_2$-regularized
least squares) with parameter $\lambda$ is defined as:
\begin{align*}
  \argmin_\w \Big\{\E_{\mu,y}\big[\big(\x^\top\w-y(\x)\big)^2\big]
  + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
  \lambda\I)^{-1}\v_{\mu,y},\quad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
\end{align*}
When Assumption \ref{a:linear} holds, then
$\v_{\mu,y}=\Sigmab_\mu\w^*$, however ridge-regularized least squares
is well-defined for much more general response models.
Our second result makes a direct connection between the (expectation
of the) unregularized minimum norm solution on the sample
and the global ridge-regularized solution.
While the under-determined regime (i.e., $n<d$) is of primary interest to us,
for completeness we state this result for arbitrary values of $n$ and $d$.
Note that, just like the definition of regularized least squares, this
theorem applies more generally than Theorem~\ref{t:mse}, in that it
does \emph{not} require the responses to follow any linear model as in
Assumption~\ref{a:linear} (proof in Appendix~\ref{s:unbiased-proof}).
\begin{theorem}[Implicit regularization of Moore-Penrose estimator]
  \label{t:unbiased}
  For $\mu$ satisfying
  % \footnote{The proof of Theorem \ref{t:unbiased}
  %   can be easily extended to probability measures $\mu$ that do not
  %   satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
  %   assumption here to simplify the presentation.}
  Assumption~\ref{a:general-position} and
  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$,
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big] =
    \begin{cases}
      (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} & \text{for }n<d,     \\
      \Sigmab_\mu^{-1}\v_{\mu,y}                 & \text{for }n \ge d,
    \end{cases}
  \end{align*}
  where, as in Theorem \ref{t:mse}, $\lambda_n$ is such that the effective dimension
  $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
\end{theorem}

@@

Our exact non-asymptotic expressions in Theorem~\ref{t:mse} and
our exact implicit regularization results in Theorem~\ref{t:unbiased}
are derived for the surrogate design, which is a
non-i.i.d.~distribution based on a determinantal point process. However, Figure \ref{f:intro}
suggests that those expressions accurately describe the MSE (up to lower order
terms) also under the standard i.i.d.~design $\X\sim\mu^n$ when $\mu$ is a multivariate Gaussian.
As a third result, we verify that the surrogate expressions for the
MSE are asymptotically consistent with the MSE of an i.i.d.~design,
for a wide class of distributions which include multivariate Gaussians.

\begin{theorem}[Asymptotic consistency of surrogate design]
  \label{t:asymptotic}
  Let $\X \in \mathbb{R}^{n \times d}$ have i.i.d.~rows $\x_i^\top = \z_i^\top\Sigmab^{\frac12}$
  where $\z_i$ has independent zero mean and unit variance
  sub-Gaussian entries, and suppose that Assumptions~\ref{a:linear}
  and \ref{a:general-position} are satisfied. Furthermore, suppose
  that there
  exist $c, C, C^* \in \mathbb{R}_{> 0}$ such that
  $C \I \succeq \Sigmab \succeq c \I \succ 0$ and $\| \w^* \| \le
    C^*$. Then
  %and the limiting spectral measure (i.e., normalized counting measure of the eigenvalues) of $\Sigmab \in \mathbb R^{d \times d}$ converges weakly to some measure $\nu$ as $d \to \infty$. Then
  \begin{align*}
    \MSE{\X^\dagger\y} - \Mc(\Sigmab, \w^*,\sigma^2,n) \to 0
  \end{align*}
  with probability one as $d,n \to \infty$ with $n/d \to \bar c \in (0,\infty) \setminus \{1\}$.
  % Let $\rho=n/d\neq 1$, $\X\sim \mu^n$
  % and $y_i=y(\x_i)$ satisfy Assumption~\ref{a:linear}. If $d\geq
  % c_\rho=\frac3{|1-\rho|}$ and
  % \begin{align*}
  %   \mu &= \begin{cases}
  %     \Nc(\zero,\I)&\text{when }n<d-1,\\
  %     \Nc(\zero,\Sigmab),\ \Sigmab\succ\zero &\text{when }n>d+1,
  %   \end{cases}
  % \end{align*}
  % then the absolute difference between surrogate
  % expressions and the true MSE is bounded as follows:
  % % &\text{if}\qquad  (1)\quad \rho<1\ \text{ and }\ \mu=\Nc(\zero,\I)\qquad\text{or}\qquad
  % %     (2)\quad \rho > 1\ \text{ and }\ \mu=\Nc(\zero,\Sigmab),\
  % %     \Sigmab\succ\zero,\\
  % \begin{align*}
  % \Big|\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)\Big|\ \leq\
  % \frac{c_\rho}{d} \cdot \Mc(\Sigmab,\w^*,\sigma^2,n).
  % \end{align*}
\end{theorem}

% \begin{remark}
% For $n$ equal to $d-1$, $d$ or $d+1$, the true MSE under Gaussian
% random design can be infinite, whereas the
% surrogate MSE is finite and has a closed form expression.
% \end{remark}

\noindent
The above result is particularly remarkable since our surrogate design
is a determinantal point process. DPPs are commonly used in ML to
ensure that the data points in a sample are well spread-out. However,
if the data distribution is sufficiently regular (e.g., a multivariate
Gaussian), then the i.i.d.~samples are already spread-out reasonably
well, so rescaling the distribution by a determinant has a negligible
effect that vanishes in the high-dimensional regime.
Furthermore, our empirical estimates (Figure~\ref{f:intro}) suggest that the surrogate expressions
are accurate not only in the asymptotic limit, but even for moderately large
dimensions. Based on a detailed empirical analysis described in
Section \ref{sec:asymp-conj-details},
we conjecture that the convergence described in
Theorem~\ref{t:asymptotic} has the rate of~$O(1/d)$.
% holds true in the under-determined regime without the assumption that
% $\Sigmab=\I$ (see Conjectures \ref{c:wishart} and
% \ref{c:projection}). In this case, no formula is known for 
% $\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
% Gaussian design naturally extend.

% \subsection{Key techniques: surrogate designs and
%   determinant preserving matrices}
